---
title: "Konza Vegetation Analysis"
author: "Owen Mees"
format: html
execute: 
  echo: false
  warning: false
  message: false
---

## Introduction

This analysis will look at vegetation data from the Konza National Ecological Observatory Network (NEON) site located near Manhhattan, KS. It will use tidyverse packages to wrangle and visualize some of the data. 

## Loading the data

The data provided in csv files. We'll use the tidyverse package `readr` to load it into R 

```{r}
#| label: read-csv

# load packages
library("readr")
library("dplyr")
library("ggplot2")
library("ggthemes")
library("knitr")

# read in the data
konza_data <- read_csv("data/NEON.D06.KONZ.DP1.10058.001.div_1m2Data.2025-08.basic.20251222T234053Z (1).csv")
```

## Data summarizing with `dplyr` 

We'll next use the dplyr package to summarize some diffeent aspects of the data. 

```{r}
#| label: summarize-with-dplyr

# first summary make will be a count of the number of observations in each family
konza_data %>% 
  group_by(family) %>% 
  tally() %>% 
  arrange(desc(n)) %>% 
  kable()
```

This analysis showed that Poaceae and Asteraceae are among the most common family found in these plots

```{r}
#| label: mean-pcnt-cover-by-family

# the goal is to calculate the mean percent cover of each family
konza_data %>% 
  group_by(family) %>% 
  summarise(mean_pcnt_cover = mean(percentCover), 
            sd_pcnt_cover = sd(percentCover)) %>% 
  arrange(desc(mean_pcnt_cover)) %>%
  kable()
```

This analysis showed that the largest percent cover is the NA taxa which means it likely wasn't able to be identified, but the highest mean percent cover at 16% across the plots was Fagaceae. 

```{r}
#| label: filter by elevation

# the goal is to calculate the mean percent cover of each family
konza_data %>% 
  filter(elevation > 390) %>% 
  group_by(family) %>% 
  summarise(mean_pcnt_cover = mean(percentCover), 
            sd_pcnt_cover = sd(percentCover)) %>% 
  arrange(desc(mean_pcnt_cover)) %>%
  kable()
```

Next we'll learn to use `mutate()`, which is a function to create a new column based on contents from an existing column. 

```{r}
#| label: use-mutate

# select pulls out the column you want
# mutate creates a new column with whatever content you want
# which can include manipulation of an existing column
konza_data %>% 
  select(scientificName, family, percentCover) %>%
  mutate(propCover = percentCover / 100) %>%
  kable()

```


Next up, we'll use `ggplot2` to make some figures!

```{r}
#| label: histogram-percent-cover
konza_data %>%
  filter(!is.na(family)) %>% 
  ggplot(aes(x = percentCover)) +
  geom_histogram() +
  labs(title = "Histogram of percent cover, all species",
       x = "Percent Cover" ) +
theme_bw() 
  


```
## Next steps 

1. Follow up on the really high percent values. Maybe say > 25%?

Looks like its mostly soil and dead plant matter. 



Need to demonstrate the use of the `mutate()` function, and then how to use `ggplot2`. Some interesting plots could be looking at the effects of elevation on plant cover

# Next steps(homework)

Need to create three figures to better understand `ggplot` and `dplyr` applications

Goal is to create a boxplot representing the cover of taxon ID by elevation. This could be used to further understand how elevation impacts taxa density. 

```{r}
#| label: taxon_id_by_elevation

konza_data %>% 
  filter(!is.na(taxonID)) %>% 
  filter(taxonID == "CODR") %>%
ggplot() +
  geom_boxplot(aes(x = elevation,
                   y = percentCover,
                   group = elevation)) +
labs(title = "Cover of taxonID by Elevation",
     y = "Percent Cover",
     x = "Elevation")
```

The boxplot shows that higher elevations tend to cause less variation in percent cover. An area to look into further could be the eleveation of 350 as it has a significantly higher percent cover of taxa. 

```{r}
#| label: elevation_histogram





```



### End homework

